Personal Activity Monitor: Steps
========================================================

The script below retrieves and analyzes the number of steps taken in five minute intervals by an anonymous individual
over the course of two months (October/November 2012).

You can retrieve the [raw data here.] (http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

In general the data consists of 288 (24 x 12 per hour)  observations each day. Unavailable data is encoded as NA. 
Intervals when the subject was not walking will show zero steps.

## Loading the data

``` {r getdata}

fname <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
tf <- tempfile()
download.file(fname,tf)

unzip(tf,"activity.csv")

raw <- read.csv("activity.csv", colClasses = c("numeric", "Date", "numeric"))
```

## Take a quick peek at the data

``` {r peek}
dim(raw)
names(raw)
missing <- tapply(is.na(raw$steps),raw$date,sum)
missing[!missing==0]
```

### Note that when data is missing for the current sample, it's missing for the whole day!

``` {r summary}
byday <- with(raw,tapply(steps,date,sum))
hist(byday, breaks=9, xlab="steps", main="Total number of Steps")

### Some summary statistics by day for reference

summary(byday)

### And similar statistics for each five minute interval

byinterval <- with(raw,tapply(steps,interval,mean,na.rm=T))

summary(byinterval)

## Graphing average activity throughout the day 

hours <- raw$interval[1:288]/100
plot(hours,byinterval, axes=F, type="l",
     xlab="time of day", ylab="number of steps",
     main="Average Daily Activity"
     )
axis(side=1,at=0:24)
axis(side=2)

whenmax <- which(byinterval==max(byinterval))
whenmax <- as.numeric(names(byinterval))[whenmax]
whenmax <- whenmax/100
```

*Maximum activity seems to occur in the five minute interval starting at `r whenmax`*

**Note that the data above is highly averaged. Compare with results for just one day below. Activity comes in tighter, more pronounced bursts.**

``` {r single, echo=FALSE}
one <- raw[raw$date=="2012-11-03",1]
plot(hours,one,axes=F,type="l",
     xlab="time of day", ylab="number of steps",
     main="Daily Activity: Nov 3, 2012"
     )
axis(side=1,at=0:24)
axis(side=2)

```

## Correcting for missing days

As noted above, data is missing for `r sum(!missing==0)` days (`r sum(is.na(raw[,1]))` observations).  

Below is an attempt to correct for missing data. Because I think that using averaged data may yield misleading results for most purposes, I have chosen to directly substitute a representative day for the missing one. Average results will likely not change much given the method chosen.

Representative days are chosen according to the following criteria:

- substitute day can't be missing data itself!
- same day of the week
- in the nearest week available

*A different strategy may be required if interday data is missing --- I would still recommend replacement over averaging results*

*Spoiler Alert: The graph at the very end of this page shows why one has to be cautious.

``` {r fixes}

dates <- as.Date(names(missing)) ## All dates in sample
dow <- weekdays(as.Date(names(missing))) ## corresponding day of the week

repone <- function(baday) {
    
    # returns the first date for which data is available
    # that matches the day of the week for the argument < baday >
    
    # checks for validity should be performed elsewhere
    # uses globals missing, dow and dates from .Rmd environment
    
    use <- missing==0 ## days without missing data
    okdow <- dow[use]; okdates <- dates[use]
    
    baday <- as.Date(baday) # just in case it comes in as character from a name field
    badow <- weekdays(baday)
    
    avail <- okdates[okdow==badow]    ## same day of week
    r <- abs(as.numeric(avail-baday)) ## how far away is it
    avail <- avail[which(r==min(r))]  ## nearest available  
    avail[1]                          ##indexed so only one day gets picked!
}

repall <- function(){
    
    ## wrapper to replace missing data
    
    out <- raw
    
    tofix <- missing[!missing==0]
    tofix <- as.Date(names(tofix))
    
    for (k in 1:length(tofix)) {
        out[out$date==tofix[k],c(1,3)] <- raw[raw$date==repone(tofix[k]),c(1,3)]
    }
    
    out
        
    }

fixed <- repall() # fix data

summary(fixed) # quick summary of fixed data
summary (raw)  # for comparison

bydayfixed <- with(fixed,tapply(steps,date,sum))
hist(bydayfixed, breaks=9, xlab="steps", main="Total number of Steps")

summary(bydayfixed)
summary(byday) # unadjusted data for comparison
```

*Note that we have been able to minimize the effects of adjusting the data. There's a slight decrease in the median; the mean remains the same.*

## Analyze weekend effects

``` {r weekend}

fixed$type <- rep("weekday",length(fixed$date))
fixed$type[weekdays(fixed$date) %in% c("Saturday","Sunday")] <- c("weekend")
fixed$type <- as.factor(fixed$type)

# decided not to make a lattice plot so I could have better control over the axes

par(mfrow=1:2)

subfixed <- fixed[fixed$type=="weekday",1:3]
byinterval <- with(subfixed,tapply(steps,interval,mean))
plot(hours,byinterval, axes=F, type="l",ylim=c(0,250),
     xlab="time of day", ylab="number of steps",
     main="Average Activity: Weekdays"
     )
axis(side=1,at=seq(0,24,6))
axis(side=2)

subfixed <- fixed[fixed$type=="weekend",1:3]
byinterval <- with(subfixed,tapply(steps,interval,mean))
plot(hours,byinterval, axes=F, type="l",ylim=c(0,250),
     xlab="time of day", ylab="number of steps",
     main="Average Activity: Weekend"
     )
axis(side=1,at=seq(0,24,6))
axis(side=2)

```

## But here's an interesting result. Look what happens when we plot the median number of steps...

```{r medianw}

par(mfrow=1:2)

subfixed <- fixed[fixed$type=="weekday",1:3]
byinterval <- with(subfixed,tapply(steps,interval,median))
plot(hours,byinterval, axes=F, type="l",ylim=c(0,120),
     xlab="time of day", ylab="number of steps",
     main="Median Activity: Weekdays"
     )
axis(side=1,at=seq(0,24,6))
axis(side=2)

subfixed <- fixed[fixed$type=="weekend",1:3]
byinterval <- with(subfixed,tapply(steps,interval,median))
plot(hours,byinterval, axes=F, type="l",ylim=c(0,120),
     xlab="time of day", ylab="number of steps",
     main="Median Activity: Weekend"
     )
axis(side=1,at=seq(0,24,6))
axis(side=2)
```

## Looks like someone has a pretty regular desk job.  
## Note the little lonely spike for lunch.